---
title: "Central Park Squirrels"
author: "Abby Sikora"
date: "2024-03-07"
categories: [news, code, analysis]
execute:
  warning: false
---
Introduction:
The data set I'm working with is from the 2018 Central Park Squirrel Consensus, which is a multimedia science, design, and storytelling project. They count squirrels and present their findings to the public. The dataset contains squirrel data for each of the 3,023 sightings, including location coordinates, age, primary and secondary fur color, elevation, activities, communications, and interactions between squirrels and with humans. I pulled this data set and information about it from the TidyTuesday github repository. Something important to note about this data is that some of the squirrels contain NA values for important variables like Age and Fur Color, so this isn't an entirely accurate colection and there are some unnamed data points being left out. 

The variables I'm most interested in are:

Age: Age of squirrel, value is either "Adult" or "Juvenile."

Primary Fur Color: Main color of their fur; value is either "Gray," "Cinnamon" or "Black."

Shift: Value is either "AM" or "PM" to communicate whether or not the sighting session occurred in the morning or late afternoon.

The questions I will be answering are:

- What is the most popular squirrel color? 
- What is the most popular color in adult and juvenile age groups respectively?
- Do adults or juveniles prefer pm over am?

```{r}
library(tidyverse)
squirrel <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-05-23/squirrel_data.csv')

#To replace NA and "?" values with Unknown for the Age and Primary Fur Color variables to make the plots look cleaner
squirrel <- squirrel |>
  mutate(`Primary Fur Color` = if_else(is.na(`Primary Fur Color`), true = "Unknown", false = `Primary Fur Color`),
         Age = if_else(is.na(Age) | Age == "?", true = "Unknown", false = Age))
```

```{r}
ggplot(data = squirrel, aes(x = `Primary Fur Color`)) +
  geom_bar(colour = "brown4", fill = "grey") +
  labs(title = "Distribution of Primary Fur Color") +
  theme_minimal()
```
When looking at this graph, we can denote that the most popular Primary Fur Color in squirrels recorded in this study is gray by a good amount (almost 2000 more squirrels are Gray than the next most popular). Cinnamon is a far second with black being a small sliver of squirrels. The unknown variable represents squirrels collected without a distinct Primary Fur Color. 

```{r}
ggplot(squirrel, aes(x = Age, fill = `Primary Fur Color`)) +
  geom_bar(data = squirrel, position = "dodge") +
  labs(title = "Distribution of Primary Fur Color by Age") +
  scale_fill_viridis_d() +
  theme_minimal()
```
By looking at this plot of Primary Fur Color by Age, we see that the distribution of fur color stays similar to the first plot I made with gray coming out on top again with Cinnamon and others far behind. Something interesting we can note without purposely seeking it out is we now see that a majority of squirrels that were recorded are Adults. 

```{r}
ggplot(squirrel, aes(x = Age, fill = Shift)) +
  geom_bar(data = squirrel, position = "dodge") +
  labs(title = "Distribution of Shift by Age") +
  scale_fill_viridis_d() +
  theme_minimal()
```
Now in this plot, we have added Shift to the conversation, looking at the sightings of squirrels by age and what time of day they were recorded. As we see, most squirrels recorded were seen during the pm, but it isn't as drastic a difference between counts for Shift than seen in previous comparisons. 

Conclusion: From the visualizations above, we can confidently conclude that after removing unidentified squirrels from the data, a typical squirrel in central park in 2018 will be a Gray Adult squirrel who is most active in the PM. 

Some flaws within the data and things I wish I had more specifics on are what were teh exact time frames that squirrels were being observed. I ask this because if a majority of squirrel observation was happening in the pm, this would obviouslt give the result we see with pm being more popular for both age groups. Another thing I wish I thought about more before using this data set is that it is mostly categorical variables, so there wasn't much I could do in terms of the usual visualizations I make. 

Future directions I would go with more data/time is exploring the variables I didn't get to look at like the different sounds the squirrels could've possibly be making, or what actiity they were doing when observed under variables like running, chasing, eating, climbing foraging etc. 

My visuals are an effective way of communicating because they are clutter free, very straight forward and use color friendly scales. 








